---
title: "Neandertal Cluster Analysis"
author: "James Bain"
date: "9/15/2017"
output: pdf_document
bibliography: biblio.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyr)
library(dplyr)
library(devtools)
library(cluster)
library(ggdendro)
library(factoextra)
library(ggfortify)
library(magrittr)
```

## Data

The data are comprised of 84 different activities compiled from a network of about 100 hospitals across the United States. This data are published in the form of Excel files and made available through the NEISS.

NEISS collects information about product and activity related injuries. Details about the anatomical region, type of injury and other demographic variables are collected. Data about injuries are then transformed to be activity specific and aggregated to match the anatomical categories defined in Berger & Trinkaus [-@berger1995]. 

```{r, include=FALSE}
# read the data
df <- read.csv('~/Documents/research/neander_trauma/Data/contingency_norm.csv',
               stringsAsFactors = FALSE)
df %<>%select(-c(`X.1`))
rownames(df) <- df$X
df %<>% select(-Xx)
```
```{r, echo = FALSE}
df[1,]
```


## Partitioning Around Medoids (PAM)

Partitioning around medoids (PAM) is a clustering algorithm that selects the

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
